---
title: "RV Builtin Rules"
output: rmarkdown::html_vignette
description: >
  The builtin RV rules and their associated Schema and Validator behaviour. 
vignette: >
  %\VignetteIndexEntry{RV Builtin Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(RV)
devtools::load_all()
```

RV has 23 validation rules and 9 cross rules.

### Validation rules

```{r builtin rules}
Registry()@rule_names
```

The builtin rules are categorised by their type: 'control', 'transform' and 'validate'. When the `Validator` is run, rules are applied in three passes according to these categories, with a special final pass for the 'apply_last' rule.

Rules within each category are applied in the order they appear in their respective `Registry` property.

```{r rule categories}
r <- Registry()
r@control_rules
r@transform_rules
r@validate_rules
```

Each validation rule has two functions associated with it: 

* A function that validates the given schema value.

```{r schema rule function example, error=TRUE}
Schema(list(type = 1L), error = TRUE)
```

* A function that uses the schema value to validate data.

```{r validation rule function example, error=TRUE}
Validator(
  data = 1L,
  schema = list(type = "character"),
  error = TRUE
)
```

### Cross rules

Cross rules operate within the `Schema`, checking that the values of two or more schema rules don't clash.

```{r cross rule example, error=TRUE}
Schema(
  list(
    min_length = 5,
    max_length = 1
  ),
  error = TRUE
)
```

## Rule information

### Validation rules

```{r, validation rules table, echo=FALSE, results='asis'}
x <- show_RV_builtins("validation")
header <- stats::setNames(rep(1, length(x)), colnames(x))
new_colnames <- as.character(x[1, ])
base_table <- stats::setNames(x[-1, ], new_colnames)

kableExtra::add_header_above(
  knitr::kable(base_table, "html", row.names = FALSE),
  header
)
```

### Cross rules

```{r, cross rules table, echo=FALSE, results='asis'}
x <- show_RV_builtins("cross")
header <- stats::setNames(rep(1, length(x)), colnames(x))
new_colnames <- as.character(x[1, ])
base_table <- stats::setNames(x[-1, ], new_colnames)

kableExtra::add_header_above(
  knitr::kable(base_table, "html", row.names = FALSE),
  header
)
```
