---
title: "Creating RV Schemas"
output: rmarkdown::html_vignette
description: >
  Create schemas that align with your data structures.
vignette: >
  %\VignetteIndexEntry{Creating RV Schemas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(RV)
```

## RV Schemas

RV schemas are nested list objects that are passed to the `Schema` class. This class automatically re-orders and validates the input list schema. The `Schema` class object is then passed to the `Validator` class for data validation. 

## Creating a schema list

The lists used for RV schemas are nested list objects with rule-named leaf elements that determine validation behaviour. The names or positions of the nested lists are used to match their rules to the corresponding data elements.

```{r schema example, eval=FALSE}
list(
  type = "data.frame",
  id = list(
    type = "numeric"
  ),
  email = list(
    type = "character",
    regex = "@gmail.com$"
  ),
  list(
    min_length = 2
  )
)
```

Rules at the top level of the nested list are applied to the whole data object. Nested list elements are matched to data elements by their name if present or by position if there is no name present. 

When matching by position, rule elements are first removed, so the first non-rule element will always be matched against `[[1]]` of the matching data node. See the following illustrations:

```{r 1 level above basic example, echo = FALSE}
lobstr::tree(list(
  "Top level rule" = "Applied to `data`.",
  list(
    "Depth 1 rule" = "Applied to `data[[1]]`.",
    list(
      "Depth 2 rule" = "Applied to `data[[1]][[1]]`."
    )
  )
))
```
```{r 1 level above basic named example, echo = FALSE}
lobstr::tree(list(
  "Top level rule" = "Applied to `data`.",
  "x" = list(
    "Depth 1 rule" = "Applied to `data[['x']]`.",
    "x" = list(
      "Depth 2 rule" = "Applied to `data[['x']][['x']]`."
    )
  )
))
```

This behaviour continues no matter the level of nesting, so it is possible to apply rules to deeply nested values.

```{r 1 level above deeply nested example, echo = FALSE}
lobstr::tree(list(list(list(list("Applied to `data[[1]][[1]][[1]]`.")))))
```

### Double hits with positional matching

When matching schema elements to data, the `Validator` first attempts to match by name before falling back to positional matching. As data elements are not flagged when validated (and thus can be validated multiple times), this can cause unexpected behaviour. See the following example where the data is matched twice:

```{r double hit example}
Validator(
  data = list(x = 1L),
  schema = list(
    list(type = "integer"), # matched positionally
    x = list(type = "character") # matched by name to same element
  )
)@errors
```

It is strongly encouraged to use fully named schemas/data unless you are certain about their structure.

## Using the Schema class

The `Schema` class takes the nested list and re-orders it, transforms certain rules from strings to functions where necessary, then validates the schema.

### Schema re-ordering

Rules are applied in three separate passes according to their category: 'control', 'transform', and 'validate', with a special rule 'apply_last' occurring in a final pass. 

For each category, the `Schema` reorders the list upon ingest according to the corresponding orders of the name properties in the `Registry` (by default `Schema` uses a default, uncustomised `Registry` if one is not provided):

```{r rule order}
r <- Registry()
r@control_rules
r@transform_rules
r@validate_rules

Schema(
  list(
    min_length = 2L,
    type = "integer",
    default = 10L,
    coerce = "double"
  )
)@schema
```

The order of the rules within each of the `Registry` properties can be edited to specify a different order, which can then be fed to the `Schema`:

```{r rule order edited}
r <- Registry()
r@validate_rules <- c("min_length", r@validate_rules[!grepl("min_length", r@validate_rules)])

Schema(
  schema = list(
    min_length = 2L,
    type = "integer",
    default = 10L,
    coerce = "double"
  ),
  registry = r
)@schema
```

For more information about the builtin rules and how they operate, or how to add custom rules to a `Registry`, see the [builtin rules vignette](RV-rules.html) and the adding rules vignette.
.

### String to function conversion

Certain rules can be given character strings as an input which are turned into functions during schema validation. The rules that this apply to can be found in the `Registry`, along with the function that does the conversion. Both can be edited.

BEWARE: No check is made on the content of the string, so use the builtin converter with extreme care for user inputs - it is vulnerable to code injection. This functionality can be removed by simply making the `@str_to_fn_rules` property an empty character.

```{r str2func example}
r <- Registry()
r@str_to_fn_rules

r@str_to_fn_converter

Schema(
  list(predicate = "function(x) x > 10")
)@schema
```

### Schema validation

`Schema` objects validate their list input and store an `@errors` property that highlights validation errors.

For each schema node it validates that:

* There are no duplicate names.
* All leaf elements are named.
* Leaf elements are name with recognised rules.

```{r schema validation}
Schema(
  list(
    x = list(type = "character"),
    x = list(type = "integer"),
    list("character"),
    list(my_rule = 1L)
  )
)@errors
```

#### Rule validation

Each rule has an associated schema validation rule that checks the value given. For example, the 'predicate' rule checks that given values are either strings or functions. The 'dependency' rule checks that given values are either a character vector (names), a numeric integerish vector, or a non-nested list containing a mix of the two.

```{r schema rule validation, error=TRUE}
Schema(
  list(
    predicate = 1L,
    dependency = 1.5
  ),
  error = TRUE
)
```

#### Cross rule validation

There are also cross rules that check if the values of multiple rules clash (if the individual rule components are themselves valid). For example the 'min_val_larger_than_max_val' rule does what it says on the tin:

```{r cross rule example, error=TRUE}
Schema(
  list(
    min_val = 5,
    max_val = 1
  ),
  error = TRUE
)
```

For full information about each rule and their validation, see `vignette('RV-rules')`.

## Key takeaways

* RV schemas are nested lists where the structure determines data matching.
* Nested lists are passed to the `Schema` class, which orders and validates a schema list.
* Each schema node applies to one level 'above' of the data.
* Rule elements are ignored when matching by position.
* Name matching takes priority over positional matching.
